Field-informed Reinforcement Learning for Learning Large-Scale Collective Tasks

### 2023-05-18 paper.tex
chore: some minors

Autonomous agents interacting and operating towards a common task, rely on shared information and consequently emerging knowledge in order to achieve the underlying goals efficiently. Coordinating and organising this information exchange within the network of agents is required for improve the collaboration among them further~\cite{jennings1996coordination}. Through this coordination, agents may overcome problems requiring (i) efficiency and accuracy only achieved by multiple agents, (ii) the complementary competencies of the individual agents, or (iii) the collective resources of multiple agents.

@@

In this paper, we propose a novel approach utilising aggregate computing~\cite{Beal2015Computer} together with a \ac{GNN}~\cite{Zhou2020AIOpen} in combination with a reinforcement learning approach, namely \ac{DQN}~\cite{mnih2015human}. 
First, aggregate computing is utilised to collect information from neighbours and create a local knowledge base for each agent. Second, we utilise a \ac{GNN} within the \ac{DQN}. Here the \ac{GNN} provides node embeddings for each agent, serving as input for the \ac{DQN}. The \ac{DQN} provides then the appropriate actions for the agent to achieve their tasks.'
%We further combine this with reinforcement learning techniques to respond to changing conditions in the environment. 
%
This follows a novel trend in which high-level declarative programming languages are mixed with machine learning techniques in order to synthesise robust collective controllers~\cite{DBLP:conf/acsos/Aguzzi21,DBLP:conf/icdcs/AguzziCV22,DBLP:conf/coordination/AguzziCV22}.
%Specifically, 
% our approach utilises aggregate computing to disseminate information by manipulating a computational field, 
% which is a distributed data structure that maps devices to information. 
% This field functions as a dynamic environment in which information is continuously updated and diffused throughout the system. 
% By leveraging this layer, we can build collective distributed intelligence using \ac{GNN},
%  which uses the information in the field to determine the best action for a given task.
%\ga{first draft, need a refinement}
The remainder of this paper is structured as follows. First, we introduce the relevant background and problem formulation in \Cref{sec:background}. Afterwards, we introduce our approach in \Cref{sec:approach}. \Cref{sec:eval} outlines the performed experiments and discusses the obtained results. Finally, we will present our conclusions in \Cref{sec:conclusion}.

@@

%
Each node is governed by an autonomous software agent, executing a \emph{local} control loop. In each iteration of this loop, the agent can access the information available within its own \emph{context}. This context is comprised by information acquired by the node directly through its sensors and the information received from its neighbours. 
%
At the end of each control loop iteration, the node can send messages to its neighbourhood. This message may contain raw sensor data or aggregated information over time or from other neighbours.
% about its current evaluation and update its internal state.
%
%This general model can be adapted to several different scenarios, such as a network of sensors, a network of robots (e.g., swarm robotics), or a network of mobile phones.
Examples of such large-scale system can be a network of robots (e.g., swarm robotics), camera and IoT networks, or a network of mobile phones.
%
Our goal is to find a \emph{homogeneously} distributed controller $\pi$ (i.e., the same controller for the whole system) that, 
 starting from \emph{only} local configurations, leads 
 the system to achieve a certain \emph{collective} requirements through \emph{cooperation}, 
 such as spatial area coverage, phenomena tracking, and node aggregation.
%
The homogeneity of the controller is a key requirement to ensure the \emph{scalability} of the approach, 
 as it allows us to avoid the need for a centralised controller that would be a bottleneck for the system,
 and it is the typical choice in swarm-like systems~\cite{brambilla2013swarm,yang2021many,pmlr-v80-yang18d,DBLP:conf/aaai/ZhengYCZZWY18}.
%
\subsection{Field-based Coordination}
%\ga{discussion about field used as a coordination mechanism in multi-agent systems. Introduction to AC}
The field-based coordination approaches utilise a concept of \emph{computational fields} 
 (or simply \emph{fields}), 
 which are \emph{distributed} data structures that associate locations with values and evolve over time. 
%
\emph{Aggregate computing}~\cite{Beal2015Computer}, a modern field-coordination approach, is rooted in earlier work on \emph{artificial potential fields}~\cite{DBLP:conf/icra/Warren89} and \emph{co-fields}~\cite{DBLP:journals/pervasive/MameiZL04}. 
% 
This \emph{macro-programming}~\cite{DBLP:journals/corr/abs-2201-03473} paradigm can be realised to devise collective and self-organizing behaviour through a composition of functions operating on fields. 
%
The fields map a set of individual agents to computational values, 
 allowing them to associate what they sense, process, and actuate. 
%
Fields are computed locally but subject to a global viewpoint, 
 enabling emergent collective behaviour through the interplay of the \emph{system model} (i.e., structure and dynamics of the systems)
 and the \emph{programming model} (i.e., how collective behaviours are expressed) known as field calculus~\cite{DBLP:conf/esocc/ViroliDB13,DBLP:journals/corr/ViroliADPB16,DBLP:journals/tocl/AudritoVDPB19}.
%
The foster is structurally similar to \emph{swarm-like systems}, 
 as it is composed of a set of \emph{agents} 
 that interact with each other through a \emph{network} of \emph{neighbourhood relations}.
%
The dynamics of each agent follow a local control loop called \emph{round}
 that comprises the following steps:
\begin{enumerate}
  \item \emph{Sense}: the agent acquires information from the environment and collects the messages of its neighbourhood, building a \emph{context};
  \item \emph{Compute}: the agent executes a \emph{program} that computes an \emph{export}, that is a set of values to be sent to the neighbourhood;
  \item \emph{Act}: the agent sends the export to its neighbourhood and updates its internal state acting on the environment.
\end{enumerate}
The proactive and iterative execution of the round by each agent leads to the emergence of collective behaviour specified by the program.

The programming model is ruled by the \emph{field calculus}~\cite{DBLP:conf/esocc/ViroliDB13,DBLP:journals/corr/ViroliADPB16,DBLP:journals/tocl/AudritoVDPB19}, 
 a core language that allows the expression of collective behaviours through the composition of \emph{functions} operating on fields.
%
Field calculus includes main operators for expressing \emph{spatio-temporal} compotations that allow for 
\emph{i)} the progression of values over time, achieved by transforming a value computed in a previous round into a new value; 
\emph{ii)} the exchange of data with neighbouring fields, where received data is validated by adjacent fields; and 
\emph{iii)} the conditional division of computation into distinct domains of collective computation. 
%
We avoid presenting here the full details of these operators since they are not essential to the contribution. 
For further elaboration on the actual field calculus, please refer to the work of~\cite{DBLP:journals/jlap/ViroliBDACP19}.

On these minimal operators, 
 it is then possible to build \emph{self-organizing} coordination blocks.
 One of the founding blocks is the \emph{gradient} essential for 
 \emph{information} flows~\cite{DBLP:conf/saso/WolfH07}.
%
This operator generates a numerical field that represents the minimum distance from a source zone. 
%
In other words, it maps a Boolean field (where ``true'' indicates the presence of a source node and ``false'' indicates its absence) to a distance field that indicates the proximity to the nearest source. In ScaFi (a scala implementation of aggregate computing) it is simply expressed as:
\begin{lstlisting}
def gradient(source: Boolean): Double
\end{lstlisting}
Along the gradient, it is possible to cast information within the system, accumulating value during the programming. This pattern is called \emph{gradient-cast} (or \emph{G}) 
\begin{lstlisting}
def G[V](source: Boolean, value: V, acc: V => V): V
\end{lstlisting}
\ga{more description here}
this is used in the use case to accumulate the distance from the phenomena that is tracked
\lukas{still not sure how the coordination works or if it has been explained in the above paragraph.}
\subsection{Graph Neural Networks}
%Over the years, several different neural networks have been proposed to solve specific tasks, ranging from 
%  simple \ac{MLP}~\cite{werbos1975beyond} to \ac{CNN}~\cite{oshea2015introduction}.
%  \lukas{more examples? }
%%
%Whereas \ac{CNN} is designed to be used with image and spatial-like data (e.g., audio), 
 \ac{GNN} is a novel neural network model used to process graph-structured data with deep learning approaches.
%
Let $G = (V, E)$ a graph  
where $E$ defines the neighbourhood relations for each participating node, 
and $V$ identifies the nodes present in the graph. %at that time.  \lukas{no time in the definition}
% 
Each node $v \in V$ is associated with an observation (or feature set) $f_v$.
For the sake of simplicity, 
 we thereafter describe $G_f$ as a graph that contains the feature set $f_v$ for each node $v \in V$.
 Note that, when we refer to $G_f$ and $G_o$ we are referring to the same graph $G$ but with different node features.
 Also, to access the feature set $f_v$ of a node $v \in V$ we use the notation $f_v$ or $G_f[v]$.
%\lukas{Over time this graph $G$ can change and therefore we define $G_t$ as the graph $G$ at a specific time $t$.} 
%
Given $f_v$, %a feature vector associated with each node $v \in V$, 
 the goal of a \ac{GNN} is to learn the node embedding $h_v$ for each node $v \in V$. The node embedding $h_v$ describes the node in the network and summarises the geometric properties of the graph in this location, allowing for comparison of various nodes in the graph.
 In modern \acp{GNN}, the node embedding $h_v$ is computed by aggregating information from the node's neighbours $\mathcal{N}_G(v)$,
  and then combining it with the node's current embedding $h_v$.
 The graph is further partitioned into layers with the node $v$ at its bottom layer $0$. With each traversal of an edge $e \in E$ from $v$, the number of layers increased
Formally, a \ac{GNN} can be defined by two phases:
\begin{equation}
a_{v}^{(k)}=\operatorname{AGGREGATE}^{(k)}\left(\left\{h_{u}^{(k-1)}: u \in \mathcal{N}_G(v)\right\}\right)   
\end{equation}

>>>>

Autonomous agents interacting and operating towards a common task, rely on shared information and consequently emerging knowledge in order to achieve the underlying goals efficiently. 
%
Coordinating and organising this information exchange within the network of agents is required for improve the collaboration among them further~\cite{jennings1996coordination}.
%
Through this coordination, agents may overcome problems requiring (i) efficiency and accuracy only achieved by multiple agents, (ii) the complementary competencies of the individual agents, or (iii) the collective resources of multiple agents.

@@

In this paper, 
we propose a novel approach utilising aggregate computing~\cite{Beal2015Computer} together with a \ac{GNN}~\cite{Zhou2020AIOpen} in combination with a reinforcement learning approach, namely \ac{DQN}~\cite{mnih2015human}. 
 First, aggregate computing is utilised to collect information from neighbours and create a local knowledge base for each agent. Second, we utilise a \ac{GNN} within the \ac{DQN}. 
 Here the \ac{GNN} provides node embeddings for each agent, serving as input for the \ac{DQN}. 
 The \ac{DQN} provides then the appropriate actions for the agent to achieve their tasks.
%We further combine this with reinforcement learning techniques to respond to changing conditions in the environment. 
%
This follows a novel trend in which high-level declarative programming languages are mixed with machine learning techniques in order to synthesise robust collective controllers~\cite{DBLP:conf/acsos/Aguzzi21,DBLP:conf/icdcs/AguzziCV22,DBLP:conf/coordination/AguzziCV22}.
%Specifically, 
% our approach utilises aggregate computing to disseminate information by manipulating a computational field, 
% which is a distributed data structure that maps devices to information. 
% This field functions as a dynamic environment in which information is continuously updated and diffused throughout the system. 
% By leveraging this layer, we can build collective distributed intelligence using \ac{GNN},
%  which uses the information in the field to determine the best action for a given task.
%\ga{first draft, need a refinement}
The remainder of this paper is structured as follows. 
 First, we introduce the relevant background and problem formulation in \Cref{sec:background}. 
 Afterwards, we introduce our approach in \Cref{sec:approach}. 
 \Cref{sec:eval} outlines the performed experiments and discusses the obtained results. 
 Finally, we will present our conclusions in \Cref{sec:conclusion}.

@@

%
Each node is governed by an autonomous software agent, 
 executing a \emph{local} control loop. 
 In each iteration of this loop, the agent can access the information available within its own \emph{context}. 
 This context is comprised of information acquired by the node directly through its sensors and the information received from its neighbours. 
%
At the end of each control loop iteration, 
 the node can send messages to its neighbourhood. 
 This message may contain raw sensor data or aggregated information over time or from other neighbours.
% about its current evaluation and update its internal state.
%
%This general model can be adapted to several different scenarios, such as a network of sensors, a network of robots (e.g., swarm robotics), or a network of mobile phones.
Examples of such large-scale systems can be a network of robots (e.g., swarm robotics), camera and IoT networks, or a network of mobile phones.
%
Our goal is to find a \emph{homogeneously} distributed controller $\pi$ (i.e., the same controller for the whole system) that, 
 starting from \emph{only} local configurations, leads 
 the system to achieve a certain \emph{collective} requirements through \emph{cooperation}, 
 such as spatial area coverage, phenomena tracking, and robot aggregation~\cite{DBLP:journals/firai/SchranzUSE20}.
%
The homogeneity of the controller is a key requirement to ensure the \emph{scalability} of the approach, 
 as it allows us to avoid the need for a centralised controller that would be a bottleneck for the system,
 and it is the typical choice in swarm-like systems~\cite{brambilla2013swarm,yang2021many,pmlr-v80-yang18d,DBLP:conf/aaai/ZhengYCZZWY18}.
%
\subsection{Field-based Coordination}
%\ga{discussion about field used as a coordination mechanism in multi-agent systems. Introduction to AC}
The field-based coordination approaches utilise a concept of \emph{computational fields} 
 (or simply \emph{fields}), 
 which are \emph{distributed} data structures that associate locations with values and evolve over time. 
%
\emph{Aggregate computing}~\cite{Beal2015Computer}, a modern field-coordination approach, is rooted in earlier work on \emph{artificial potential fields}~\cite{DBLP:conf/icra/Warren89} and \emph{co-fields}~\cite{DBLP:journals/pervasive/MameiZL04}. 
% 
This \emph{macro-programming}~\cite{DBLP:journals/corr/abs-2201-03473} paradigm can be realised to devise collective and self-organizing behaviour through a composition of functions operating on fields. 
%
The fields map a set of individual agents to computational values, 
 allowing them to associate what they sense, process, and actuate. 
%
Fields are computed locally but subject to a global viewpoint, 
 enabling emergent collective behaviour through the interplay of the \emph{system model} (i.e., structure and dynamics of the systems)
 and the \emph{programming model} (i.e., how collective behaviours are expressed) known as field calculus~\cite{DBLP:conf/esocc/ViroliDB13,DBLP:journals/corr/ViroliADPB16,DBLP:journals/tocl/AudritoVDPB19}.
%
The system model is structurally similar to \emph{swarm-like systems}, 
 as it is composed of a set of \emph{agents} 
 that interact with each other through a \emph{network} of \emph{neighbourhood relations}.
%
The dynamics of each agent follow a local control loop called \emph{round}
 that comprises the following steps:
\begin{enumerate}
  \item \emph{Sense}: the agent acquires information from the environment and collects the messages of its neighbourhood, building a \emph{context};
  \item \emph{Compute}: the agent executes a \emph{program} that computes an \emph{export}, that is a set of values to be sent to the neighbourhood;
  \item \emph{Act}: the agent sends the export to its neighbourhood and updates its internal state acting on the environment.
\end{enumerate}
The proactive and iterative execution of the round by each agent leads to the emergence of collective behaviour specified by the program.

The programming model is ruled by the \emph{field calculus}~\cite{DBLP:conf/esocc/ViroliDB13,DBLP:journals/corr/ViroliADPB16,DBLP:journals/tocl/AudritoVDPB19}, 
 a core language that allows the expression of collective behaviours through the composition of \emph{functions} operating on fields.
%
Field calculus includes main operators for expressing \emph{spatio-temporal} compotations that allow for 
\emph{i)} the progression of values over time, achieved by transforming a field computed in a previous round into a new field; 
\emph{ii)} the exchange of data with neighbouring fields, where received data is accessed by adjacent fields; and 
\emph{iii)} the conditional division of computation into distinct domains of collective computation. 
%
%We avoid presenting here the full details of these operators since they are not essential to the contribution. 
For further elaboration on the actual field calculus, please refer to the work of~\cite{DBLP:journals/jlap/ViroliBDACP19}.

On these minimal operators, 
 it is then possible to build \emph{self-organizing} coordination blocks.
 One of the founding blocks is the \emph{gradient} essential for 
 \emph{information} flows~\cite{DBLP:conf/saso/WolfH07}.
%
This operator generates a numerical field that represents the minimum distance from a source zone. 
%
In other words, it maps a Boolean field (where ``true'' indicates the presence of a source node and ``false'' indicates its absence) to a distance field that indicates the proximity to the nearest source. 
%
In ScaFi~\cite{casadei2022scafi} (a scala implementation of aggregate computing) it is simply expressed as:
\begin{lstlisting}
def gradient(source: Boolean): Double
\end{lstlisting}
Along the gradient, it is possible to cast information within the system, accumulating value during the field expansion. 
This pattern is called \emph{gradient-cast} (or \emph{G}) and is a fundamental building block for the coordination of swarm-like systems.
It is expressed as: 
\begin{lstlisting}
def G[V](source: Boolean, value: V, acc: V => V): V
\end{lstlisting}
%\ga{more description here}
where \texttt{source} is a Boolean field that indicates the presence of a source node, 
 \texttt{value} is the value to be cast, and 
 \texttt{acc} is the accumulation function that is applied to the value during the cast.

\lukas{still not sure how the coordination works or if it has been explained in the above paragraph.}
\subsection{Graph Neural Networks}
%Over the years, several different neural networks have been proposed to solve specific tasks, ranging from 
%  simple \ac{MLP}~\cite{werbos1975beyond} to \ac{CNN}~\cite{oshea2015introduction}.
%  \lukas{more examples? }
%%
%Whereas \ac{CNN} is designed to be used with image and spatial-like data (e.g., audio), 
 \ac{GNN} is a novel neural network model used to process graph-structured data with deep learning approaches.
%
Let $G = (V, E)$ a graph  
where $E$ defines the neighbourhood relations for each participating node, 
and $V$ identifies the nodes present in the graph. %at that time.  \lukas{no time in the definition}
% 
Each node $v \in V$ is associated with an observation (or feature set) $f_v$.
For the sake of simplicity, 
 we thereafter describe $G_f$ as a graph that contains the feature set $f_v$ for each node $v \in V$.
 Note that, when we refer to $G_f$ and $G_o$ we are referring to the same graph $G$ but with different node features.
 Also, to access the feature set $f_v$ of a node $v \in V$ we use the notation $f_v$ or $G_f[v]$.
%\lukas{Over time this graph $G$ can change and therefore we define $G_t$ as the graph $G$ at a specific time $t$.} 
%
Given $f_v$, %a feature vector associated with each node $v \in V$, 
 the goal of a \ac{GNN} is to learn the node embedding $h_v$ for each node $v \in V$. The node embedding $h_v$ describes the node in the network and summarises the geometric properties of the graph in this location, allowing for comparison of various nodes in the graph.
 In modern \acp{GNN}, the node embedding $h_v$ is computed by aggregating information from the node's neighbours $\mathcal{N}_G(v)$,
  and then combining it with the node's current embedding $h_v$.
 The graph is further partitioned into layers with the node $v$ at its bottom layer $0$. With each traversal of an edge $e \in E$ from $v$, the number of layers increased
Formally, a \ac{GNN} can be defined by two phases:
\begin{equation}
a_{v}^{(k)}=\operatorname{AGGREGATE}^{(k)}\left(\left\{h_{u}^{(k-1)}: u \in \mathcal{N}_G(v)\right\}\right)   
\end{equation}
